---
title: "In Class Practice - Lecture 2 - Sep 3"
author: "Phuong Vo"
date: today
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
---

# Learning Objectives

In this practice you will do:

1.  Create a Quarto document.
2.  Run R code chunks inside Quarto.
3.  Write mathematical equations using LaTeX.
4.  Render a Quarto document.
5.  Identify files created during rendering.
6.  Set a working directory in R.
7.  Load a dataset from a package.
8.  Save and reload data using `.rda` files.
9.  Basic practices with vectors and lists.

# Creating a Quarto Document

A Quarto document combines:

- Text
- R code
- Figures
- Mathematical equations
- Output reports

## A Simple Code Chunk

The code below calculates the mean of a sequence of numbers.

```{r}
x <- 1:10
sum(x)
```

Use the chunk options to control exactly what you want to see in the file output documents. For example, echo: whether showing the code itself, eval: whether executing the code, include: whether anything from the chunk appears.

```{r}
m = mean(x)
```

We can also print out the values by calling it out in text. For example, the mean is `r m`.

## Creating a Plot

```{r}
plot(x, x^2,
     main = "Simple Example Plot",
     xlab = "x",
     ylab = "x squared")
```

## Using LaTeX

Quarto supports LaTeX mathematics.

The quadratic formula is

$$y = \beta_0 + \beta_1 x + \epsilon$$

Another example:

$$\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$$

# Rendering the Document

Click the **Render** button in RStudio.

Or run:

```{r eval=FALSE}
quarto::quarto_render()
```

# Files Generated by Rendering

Suppose our source file is:

``` text
quarto_test.qmd
```

After rendering to HTML, you will typically see:

``` text
quarto_test.qmd
quarto_test.html
quarto_test_files/
```

The `_files` folder stores:

- Figures
- CSS
- JavaScript
- Other supporting files

# Exercise 1

## Create your own Quarto document.

Requirements:

1.  Change the title.
2.  Add a code chunk that calculates a summary statistic.
3.  Add one mathematical equation using LaTeX.
4.  Render the document.
5.  Examine all files created after rendering.

## Check Your Working Directory

```{r}
getwd()
```

## Set Your Working Directory

Replace the path below with your own folder.

```{r}
setwd("/Users/phvo/Desktop/STATS506_inclass/sep3/data")
```

## Load the Package

If needed:

```{r eval=FALSE}
install.packages("lme4")
```

Load the package:

```{r}
library(lme4)
```

# Exercise 2: Work with data

In your created quarto document above, complete the following tasks:

1.  Set your working directory.
2.  Load the `lme4` package.
3.  Load the `InstEval` dataset.

```{r}
data = lme4::InstEval
head(data)
```


4.  Display:
    - `head()`
    - `str()`
    - `summary()`
    
```{r}
summary(data)
```

5.  Save the dataset as an RDA File:

``` text
InstEval_09_03_2026.rda
```

```{r}
saveRDS(data, file = "data/InstEval_09_03_2026.rda")
```


6.  Verify the file appears in your specified folder.

# Exercise 3: Work with vectors.

1.  Create `x` as a vector of 10 elements with random values.
```{r}
x <- sample(1:20, 10)
```

2.  Extract the value at the 5th position.

```{r}
x[5]
```

3.  Extract the values from index 2 to index 5.

```{r}
x[2:5]
```

4.  Remove (drop) the values from index 2 to index 5 from the vector.

```{r}
x[-(2:5)]
```

5.  Print all values in the vector that are positive.

```{r}
x[x>10]
```

6.  Using the above vector, try the following code

```{r}
#| eval: false
x[-2:5]
```

7.  Name all of values in the vector you created in 4. For example, `room <- c("linda" = 1, "bob" = 6, "mike" = 7, "alex" = 8, "peter" = 9, "PJ" = 10)`.

```{r}
room <- c("linda" = 15, "bob" = 14, "mike" = 13, "alex" = 3, "peter" = 16, "PJ" = 19)
```

# Exercise 4: Work with lists.

1.  Create a list as same as the named vector created in 7. Check your result with the function `is.list()`.

```{r}
room_list <- list("linda" = 15, "bob" = 14, "mike" = 13, "alex" = 3, "peter" = 16, "PJ" = 19)
```

2.  Subset the first object in your list.

```{r}
room_list$linda
```

3.  Subset the value of the first object in your list.

```{r}
room_list[[1]]
```

4.  Print out all attributes of your list.

```{r}
is.list(room)
```

5.  Extract one attribute of your list.

```{r}
attributes(room_list)
```

6.  Add a new attribute to your list. For example, I would add "department" into my list.

```{r}
room_list_2 = list(room)
```

7.  Assign NULL to remove the attribute you just added in #6.


