---
title: "In Class Practice - Lecture 3 - Sep 8"
author: "Phuong Vo"
date: today
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
---

# Learning Objectives

In this practice you will do:

1.  Intro Git/ Github.
2.  Git Preliminary Assignment:
    1.  Create a repository on your Github (stat506_test_repo). Add SSH key.
    2.  Clone the repository to your local computer using SSH path.
    3.  Make any changes in these files: `.gitignore`, `README.md`
    4.  Stage, commit, push those changes to the repository on Github.
    5.  On Github site, make some more changes in the repository.
    6.  Stage, commit, push those changes to the repo on your local computer.
3.  Advanced topics: branches, rebasing, blame.
4.  Continue basic practices in R (matrix, array, data frame, logicals, functions)
5.  Announcement: Today (Sep 08) office hour is online only. Please schedule your meeting in advance using [this booking page](https://calendar.app.google/zyyyko6DQqUoRkD69).

# Exercise 1: Repository for Homework 1.

Create a GitHub repository, clone it locally, create a Quarto document, and include the repository link inside the document.

For example: In your "Homework_1" repository folder, create a file named "hw1.qmd" with the following content:

"The link to the GitHub repository of this assignment is:

https://github.com/phuongov/Homework_1."

# Exercise 2: Work with lists.

1.  Create a list of names and room assignment. Check your result with the function `is.list()`.

```{r}
room_list <- list(room_assignment = sample(1:20, 6), 
                  employee = c("linda",
                                "bob",
                                "mike",
                                "alex",
                                "peter",
                                "PJ")
                  )

room_list

is.list(room_list)
```

2.  Subset the first object in your list.

```{r}
room_list$room_assignment

# or
room_list[["room_assignment"]]
```

3.  Subset the value of the first object in your list.

```{r}
room_list$room_assignment[2:5]
```

4.  Print out all attributes of your list.

```{r}
attributes(room_list)
```

Every named list stores its element names as an attribute. In fact, `attributes(room_list)$names` is equivalent to `names(room_list)`

You can see the structure of a list by `str()`.

5.  Extract one attribute of your list.

```{r}
attr(room_list, "names")
```

6.  Add a new attribute to your list. For example, I would add "department" into my list.

```{r}
attr(room_list, "department") = c("STATS")
str(room_list)
```

7.  Assign NULL to remove the attribute you just added in #6.

```{r}
attr(room_list, "department") = NULL
str(room_list)
```

# Exercise 3: Work with other classes of an object.

## Matrix and array

1.  Matrix class: Create an matrix 2 by 6 with values are numbers from 1 to 12.

```{r}
mt = matrix(1:12, 2, 6)
```

2.  Transpose the matrix above

```{r}
dim(mt) = c(6,2)
```

3.  Split the matrix into 2 array of dimension (3 x 2)

```{r}
dim(mt) = c(3,2,3)
```
4.  Extract the values at 2nd row, 1st collumn of the 2nd array

```{r}
mt[2, 1, 2]
```

5.  Extract the entire 2nd array as x2.

```{r}
x2 = mt[ , , 2]
x2
```

```{r}
x3 = mt[ , , 2, drop = FALSE]
x3
```


## Data frame

1.  Create a data frame of 10 rows by 5 columns which are “id”, “group”, “var1”, “var2”, “var3”.

```{r}
mydat <- data.frame(
  id = 1:10,
  group = c("A", "A", "A", "A", "A",
            "B", "B", "B", "B", "B"),
  var1 = c(10, 12, 15, 18, 20, 22, 25, 28, 30, 35),
  var2 = c(100, 120, 110, 130, 140, 150, 145, 160, 170, 180),
  var3 = rep(c("stat506-001", "stats506-002"), 5)
)

mydat
```

2.  Extract the first column “id”. Extract multiple columns.
3.  Extract the first row. Extract rows for Group A.
